\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-12-07 paper.tex
improved relation to FRP
processed final review meeting’s remarks

\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien, Danny Hughes, 
Christophe Huygens, Wouter Joosen}%
\IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
3001 Leuven, Belgium\\
\{firstname.lastname\}@cs.kuleuven.be}}

@@

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns and anomalies that can indicate malicious
behavior. However, supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constrained devices such as those in
the Internet of Things. Fusion of the algorithms reduces resource consumption
by optimizing resource sharing, yet, proves to be time-consuming, repetitive
and error-prone. To address this problem we propose \NAME, a framework for the
development of efficient intrusion detection systems. It consists of a domain
specific language, to formally describe intrusion detection algorithms on a
functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that \NAME-generated code is
smaller and reduces message passing overhead, execution time and memory
footprint in comparison to a straightforward sequential implementation of the
same algorithms.

@@

@@

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health and enhanced
learning \cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes and personal data to
all of these interconnected devices, we also open them to everyone who is able
to break the virtual locks that protect them (i.e.\ passwords, encryption
keys). Implementation of security measures by itself is hindered by the fact
that IoT devices typically have limited batteries, processing power and memory.

% classic networks: fw and central ids

In classic networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection system (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerts the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

@@

In wireless networks of resource-constrained devices, which make up a
significant part of the IoT, it is not possible for a single point in the
network to oversee all traffic \cite{mishra2004intrusion}. Every device has to
implement its own lines of defense. But making the IDS a local service on each
device requires local resources, which are a scarce commodity for IoT devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, through the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service. For each of these attacks an algorithm needs to look for patterns
or anomalies. An IDS for IoT devices therefore requires a large number of
implemented algorithms for an adequate coverage.

@@

Current developments in ID on resource-constrained devices focus on
programming frameworks that structure the implementation of algorithms
\cite{valero2012di} and offer the required basic functional components to
implement them \cite{krontiris2008lidea}. A key shortcoming of prior ID
frameworks, is that they don't offer a way to optimize the usage of resources.
A solution that supports the integration of multiple algorithms on IoT devices
needs to avoid accumulating their impact on the available resources. A second
omission is support for systematic reuse to create different configurations
heterogeneous devices. Addressing this in a transparent and automated way
offers the potential to significantly reduce the effort of developing an IDS.

@@

% 1. framework: language + code generator

The first contribution of this paper is a framework called \NAME. It enables
the formal description of ID algorithms on a functional level and provides a
code generator to produce corresponding source code for a given platform and
configuration. This way, \NAME addresses the heterogeneous nature of both the
IoT and ID algorithms.

% 2. FCF to optimise

The second contribution introduces functional code fusion (FCF) as a source
code generation paradigm to solve the problem of hard to reuse and combine ID
algorithms. FCF identifies common data and functions, and organizes code to
eliminate redundant iterations, tests and computations. Results show reduced
usage of the wireless radio, reduced execution time, lower memory usage and
improved image size.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
analyses and classifies ID algorithms. Section \ref{pattern} looks for patterns
in the implementation of these classes. Section \ref{problems} identifies the
problems arising from these patterns. Section \ref{design} describes the design
we applied to construct \NAME, its domain specific language (DSL) and code
generator. Section \ref{evaluation} discusses and evaluates an implementation.
Section \ref{related} explores related work in the field of DSLs and code
generation. Finally, section \ref{conclusion} summarizes our findings, draws
conclusions and identifies topics for future work.

@@

Two aspects, currently missing from ID frameworks, that \NAME wants to address,
are optimization of resources and support for systematic reuse. This section
uses a classification to define the scope of \NAME, detailing which types of
algorithms are good candidates for the proposed approach.

@@

Because not all devices are always actively participating in the network, they
can miss events that would otherwise lead them to detect intrusions. It is
hardly impossible for a single device to detect a complex attack by itself.
Therefore a second important research topic consists of cooperative algorithms
to combine information from single identities into a group-based decision about
alleged intrusions. We identify two ways to construct cooperative decision
making and focus on the way participants exchange information: broadcast and
interactive communication.

@@

  \item[Broadcast communication] can allow implementing a distributed,
  cooperative algorithm locally: devices can exchange information about the
  reputation of other devices, solely using one-way broadcast messages.
  Combining this information allows them to decide about their trust in a given
  device, based on more than their own, often partial, observations. The
  reputation-based detection algorithm introduced in
  \cite{ganeriwal2008reputation} illustrates this. It uses the same principle
  of a watchdog also found in \cite{mishra2004intrusion}.

@@

Not all algorithms can decide if an attack is real or not, based on a single
event. Signature and misuse detection can in some cases, but anomaly detection
typically deals with some level of uncertainty. Also, in distributed situations
it is impossible to fully trust any neighboring device. In these situations one
has to rely on statistics and thresholds to decide when to drop trust in a
given device.

@@

The pattern calls the different, linked implementations in sequence in two
situations: first, when the device observes new messages on the network and,
secondly, at a regular interval to allow for internal housekeeping, such as
threshold evaluation.

These patterns can now help in identifying the problem areas that tend to
result in suboptimal code when combining multiple ID algorithms. Besides these
code-level problems, operational problems also hinder the implementation of an
IDS for IoT devices.

Two issues arise from the nesting of loops that originate from the combination
of the ID algorithm pattern and the application pattern: loops and common data.
A third issue is the use of the network.

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can incur overhead. An example from
  the ID algorithm pattern is the handling of messages: each algorithm parses
  each observed message on the network, processing the same byte stream,
  performing comparable operations to find similar patterns.

  \item[The network] allows algorithms to operate on a distributed scale. But
  scattered calls to the \ttt{SEND} function cause multiple accesses to the
  wireless radio, keeping it active and energy consuming more than strictly
  needed.

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This also explains why there
  are few implementations readily available to developers: there is no single
  platform to target for researchers.

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms.
But at the same time they conflict with their target environment, the IoT\@.
\NAME provides a framework to efficiently implement an IDS for the IoT and
offers a solution for the identified problems. It consists of a 1) DSL, 2)
source code generator and 3) software library. Figure \ref{fig:design}
visualizes the overall design.

@@

A DSL and code generator defines a development strategy. This has implications,
both on what is possible and how it can be achieved. We will first position
\NAME and motivate its design with respect to these trade offs. Second, we will
look at the principal design components of the DSL. Third, we discuss the
structure of the code generator and introduce its software library, that offers
a way to deal with for example the network access issue.

@@

The applied grayscale indicates the components' platform dependency: the DSL is
platform and language independent and deals with subsystems, such as the
network, in an abstract way. The Nodes domain implements some of these
abstracted subsystems, but still in a platform and language independent way.
The platform and language descriptions define the target platform and language
and convert the abstract constructions in actually supported code. The software
library is standard native code, targeting the platform and language. Finally,
the code and image are artifacts of the automated generation and compilation
process. They aren't a product of manual work and therefore aren't directly
dependent on the platform.

This hierarchy is important from a development effort point of view. The level
of dependency on the targeted platform also dictates the level of effort to
implement changes to that platform. For example: to generate a different
language will require a complete rewrite of the software library and a partial
rewrite of the language description of the generator. Although this is a
reusable, one-time effort, it is important in relation to the scalability of
the framework itself.

@@

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \NAME's DSL and model focuses on avoiding constructions
that result in problems concerning loops, common data and the use of the
network. To do this, \NAME uses an event-based approach to defining
functionality.

@@

Most, if not all, ID algorithms deal with information about other devices in
the network. To centralize this information \NAME provides a unified
\ttt{nodes} concept as part of its DSL and exposes it as an object to the
implementer.

Algorithms can extend nodes with their own properties. This way, only one
instance for every known device will be present in memory. Finally, nodes can
define a scope for a function to execute in. This way, nodes also offer a way
to i.e.\ iterate all known devices, again without the need for an explicit loop.

@@

Using a platform independent language also offers the possibility to relieve
the implementer of explicit memory management and low-level data structures.
From a language perspective, \NAME offers several programming constructs,
typically found in modern programming languages with higher levels of
abstraction: optional typing, object orientation, pattern matching, aspect
orientation.

@@

To convert the DSL-based ID algorithms into production-ready code, \NAME
applies code generation techniques. Although different approaches here are
applicable, like template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the required transformations demand a
richer model.

@@

The syntactic model is an AST for a virtual, expressive language. It contains
constructs and ideas found in different languages and programming paradigms. No
single existing programming language supports all features as is. Syntactic
transformations therefore morph the unsupported constructs into comparable
constructs supported by the targeted platform and language.

@@

Other common functionality also isn't generated as is. The execution of
functions based on a schedule or the implementation of higher level conceptual
functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
software library that supports the code generator.

Optimization of such algorithms is not within the scope of \NAME. It also gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

@@

The DSL is also platform and language independent. This allows the source code
generator to target different configurations without any change to the formal
descriptions of the algorithms.

@@

In this section we present a prototype implementation and evaluate its
performance. We first present the implementation consisting of the hardware
platform, network and software stack. Next we use this implementation to
quantify its merits using a side-by-side comparison of generated code versus a
manual implementation. We consider experimental results regarding usage of the
network, execution time of the event-loop and the image size. We also take a
more analytical approach and look at the handling of common data and the size
of the source code.

@@

@@

We evaluated four criteria: the network usage in number of packets and bytes,
the time required to perform once cycle of the event loop and the image size of
the resulting compiled code. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

@@

Secondly, when turning to the goal to reduce the network usage, we see that
both the number of packets as well as the amount of bytes sent have dropped by
13\% and 7\% respectively.

@@

Looking at the generated source code is another way to validate the framework.
A simple metric is comparing the amount of code versus the size of the formal
description of ID algorithms. This can be an indication of the development
effort and complexity. In this experiment, implementing both algorithms
manually results in 637 lines of C code. The corresponding \NAME implementation
results in 156 lines of DSL code. This is largely thanks to the higher level
coding constructs, as introduced in section \ref{language-support} dealing with
\NAME's language support.

We can also look at data structures. Sharing data eliminates redundant
information, such as a network address, from the scope of each individual
algorithm. Analysis of the source code with both algorithms implemented, shows
that this reduces the memory footprint for one node entry by 2 bytes, from 27
bytes to 25 bytes. A ZigBee network address indeed consists of 16 bits. So this
results in a reduction of 7.4\%. Using ZigBee's 64 bits unique addresses, this
reduction can easily become 20.5\%.

@@

Adding multiple intrusion detection algorithms to wireless resource-constrained
devices requires fusion of algorithms on a functional level. This can reduce
execution inefficiencies, redundant memory usage and superfluous network access.

@@

As a next step, we want to investigate the relation of \NAME to other
programming paradigms such as functional reactive programming. They share
common concepts with respect to event-drivenness and the preference to avoid
explicit loops.

@@

This research is partially funded by the Interuniversity Attraction Poles
Programme Belgian State, Belgian Science Policy, and by the Research Fund KU
Leuven. This research is partially funded by the EU FP7 project NESSoS\@. We
would like to thank the reviewers for their thoughtful and helpful comments
that enhanced the readability of this paper.

>>>>

\author{\IEEEauthorblockN{Christophe Van Ginneken, Jef Maerien,
Christophe Huygens, Wouter Joosen, Danny Hughes}%
\IEEEauthorblockA{iMinds-DistriNet, KU Leuven\\
3001 Leuven, Belgium\\
\{firstname.lastname\}@cs.kuleuven.be}}

@@

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns and anomalies that can indicate malicious
behavior. However, supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constrained devices such as those in
the Internet of Things. Manual fusion of the algorithms reduces resource
consumption by optimizing resource sharing, yet, proves to be time-consuming,
repetitive and error-prone. To address this problem we propose \NAME, a
framework for the development of efficient intrusion detection systems. It
consists of a domain specific language and a code generator. The language
allows formally describing the intent of intrusion detection algorithms and
supports the generator in organizing the source code. A side-by-side comparison
shows that \NAME-generated code reduces message passing overhead, execution
time and memory footprint in comparison to sequential calls to individual
implementations of the algorithms.

@@

% wired networks: fw and central ids

In wired networks, firewalls focus on the outer perimeter of the network,
filtering unwanted packets and protecting the entire internal network. But
attacks on flaws in services can pass unnoticed. This is where intrusion
detection (ID) comes into play: an intrusion detection system (IDS) monitors
all traffic that passes through the firewall, looking for patterns of malicious
activity, and optionally, after detecting such pattern, alerts the firewall,
allowing it to take corrective actions \cite{denning1987intrusion}.

@@

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health and enhanced
learning \cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes and personal data to
all of these interconnected devices, we also open them to everyone who is able
to break the virtual locks that protect them (i.e.\ passwords \& encryption
keys). Implementation of security measures by itself is hindered by the fact
that IoT devices typically have limited batteries, processing power and memory.

@@

In wireless networks of resource-constrained devices, which make up a
significant part of the IoT, it is not possible for a single point in the
network to oversee all traffic \cite{mishra2004intrusion}. Every device has to
implement its own lines of defense. However making the IDS a local service on
each device requires local resources, which are a scarce commodity for IoT
devices.

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the physical layer, through the
access and routing layers, up to the application layer. Each layer presents
different opportunities to manipulate data, eavesdrop or perform a form of
denial of service. For each of these attacks an algorithm needs to look for
patterns or anomalies. An IDS for IoT devices therefore requires a large number
of algorithms for an adequate coverage.

@@

Current developments in ID on resource-constrained devices focus on programming
frameworks that structure the implementation of algorithms \cite{valero2012di}
and offer the required basic functional components to implement them
\cite{krontiris2008lidea}. A key shortcoming of prior ID frameworks, is that
they don't offer a way to optimize the usage of resources. A solution that
supports the integration of multiple algorithms on IoT devices needs to avoid
accumulating their impact on the available resources. A second omission is
support for systematic reuse to create different configurations. Addressing
this in a transparent and automated way offers the potential to significantly
reduce the effort of developing an IDS.

@@

% 1. classification of ID and design pattern for ID algorithms

The first contribution of this paper consists of the formulation of a design
pattern for ID algorithms, based on properties identified from a classification.

% 2. framework: language + code generator

The second contribution is a framework called \NAME. It enables the formal
description of ID algorithms on a functional level and provides a code
generator producing source code for a given platform and configuration. This
way, \NAME addresses the heterogeneous nature of both IoT and ID algorithms.

% 3. FCF to optimise

The third contribution introduces functional code fusion (FCF) as a source code
generation paradigm to address hard to combine ID algorithms. FCF identifies
common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show reduced usage of the wireless
radio, execution time and memory usage.

@@

The remainder of this paper proceeds as follows: section \ref{problems}
describes the inherent problems with the combination of ID algorithms. Section
\ref{classification} analyses and classifies ID algorithms. Section
\ref{pattern} looks for patterns in the implementation of these classes.
Section \ref{design} describes the design we applied to construct \NAME, its
domain specific language (DSL) and code generator. Section \ref{evaluation}
discusses and evaluates an implementation. Section \ref{related} explores
related work in the field of DSLs and code generation. Finally, section
\ref{conclusion} summarizes our findings, draws conclusions and identifies
topics for future work.

The combination of multiple ID algorithms tends to result in suboptimal code.
Besides these code-level problems, operational problems also hinder the
implementation of an IDS for IoT devices.

Two issues arise from the way ID algorithms are implemented and deal with loop
constructs and common data. A third issue is the use of the network.

  \item[Loops] by themselves are not a problem, but different loops, that
  perform the same operations in sequence, can incur overhead. An example is
  the handling of messages: each algorithm parses each observed message on the
  network, processing the same byte stream, performing comparable operations to
  find similar patterns.

  \item[The network] allows algorithms to operate on a distributed scale. But
  scattered calls to the a message sending function cause multiple accesses to
  the wireless radio, keeping it active and energy consuming more than strictly
  needed.

  \item[The heterogeneity of IoT devices] results in different software stacks,
  each with for example their own network API\@. This partially explains why
  there are few implementations readily available to developers: there is no
  single platform to target for researchers.

@@

Two aspects currently missing from ID frameworks are optimization of resources
and support for systematic reuse. This section uses a classification to define
the scope of \NAME, detailing which types of algorithms are good candidates for
the proposed approach.

@@

Because not all devices are always actively participating in the network, they
can miss events that would otherwise lead them to detect intrusions. It is
hardly impossible for a single device to detect a complex attack by itself.
Therefore a second important class of ID algorithms consists of cooperative
algorithms that combine information from single identities into a group-based
decision about alleged intrusions. We identify two ways to construct
cooperative decision making and focus on the way participants exchange
information: broadcast and interactive communication.

@@

  \item[Broadcast communication] can allow implementing a distributed,
  cooperative algorithm locally: devices can exchange information about the
  reputation of other devices, solely using one-way broadcast messages.
  Combining this information allows them to decide about their trust in a given
  device, based on more than their own, often partial, observations. The
  reputation-based detection algorithm introduced in
  \cite{ganeriwal2008reputation} illustrates this. It uses the same principle
  of a watchdog found in \cite{mishra2004intrusion}.

@@

Not all algorithms can decide if an attack is real or not, based on a single
event. Signature and misuse detection can in some cases, but anomaly detection
typically deals with some level of uncertainty. Also, in distributed setups it
is impossible to fully trust any neighboring device. In these situations one
has to rely on statistics and thresholds to decide when to drop trust in a
given device.

@@

The pattern calls the individual, linked implementations in sequence in two
situations: first, when the device observes new messages on the network and,
secondly, at a regular interval to allow for internal housekeeping, such as
threshold evaluation.

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms.
But at the same time they conflict with the target environment, the IoT\@.
\NAME provides a framework to efficiently implement an IDS for the IoT and
offers a solution for the identified problems. It consists of a 1) DSL, 2)
source code generator and 3) software library. Figure \ref{fig:design}
visualizes the overall design.

@@

Using a DSL and code generator defines a development strategy. This has
implications, both on what is possible and how it can be achieved. We will
first position \NAME and motivate its design with respect to these trade offs.
Second, we will look at the principal design components of the DSL. Third, we
discuss the structure of the code generator and introduce its software library,
that offers a way to deal with for example the network access issue.

@@

The applied grayscale indicates the components' platform dependency: the DSL is
platform and language independent and deals with subsystems, such as the
network, in an abstract way. The Nodes domain implements some of these
abstracted subsystems, but still in a platform and language independent way.
Platform and language descriptions convert the abstract constructions in
actually supported code. The software library is standard native code,
targeting the platform and language. Finally, the code and image are artifacts
of the automated generation and compilation process. They aren't a product of
manual work and therefore aren't directly dependent on the platform.

This hierarchy is important from a development effort point of view. The level
of dependency on the targeted platform dictates the level of effort to
implement changes to that platform. For example: to generate a different
language will require a complete rewrite of the software library and a partial
rewrite of the language description of the generator. Although this is a
reusable, one-time effort, it is important in relation to the scalability of
the framework itself.

@@

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \NAME's DSL and model focus on avoiding constructions
that result in problems concerning loops, common data and the use of the
network. To do this, \NAME uses an event-based approach to defining
functionality.

@@

Most, if not all, ID algorithms deal with information about other devices in
the network. To share this information \NAME provides a unified \ttt{nodes}
concept as part of its DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This way, only one
instance for every known device will be present in memory. Finally, nodes can
define a scope for a function to execute in. This way, nodes offer a way to
i.e.\ iterate all known devices, again without the need for an explicit loop.

@@

Using a platform independent language offers the possibility to relieve the
implementer of explicit memory management and low-level data structures. From a
language perspective, \NAME offers several programming constructs, typically
found in modern programming languages with higher levels of abstraction:
optional typing, object orientation, pattern matching, aspect orientation.

@@

To convert the DSL-based ID algorithms into production-ready code, \NAME
applies code generation techniques. Although different approaches are
applicable, like template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the required transformations demand a
richer model.

@@

The syntactic model is like an abstract syntax tree (AST) for a virtual,
expressive language. It contains constructs and ideas found in different
languages and programming paradigms. No single existing programming language
supports all features as is. Syntactic transformations therefore translate the
unsupported constructs into comparable constructs supported by the targeted
platform and language.

@@

Other common functionality isn't generated as is either. The execution of
functions based on a schedule or the implementation of higher level conceptual
functions, such as \ttt{now} and \ttt{sha1}, are calls into the standard
software library.

Optimization of such algorithms is not within the scope of \NAME. This gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

@@

The DSL is platform and language independent. This allows the source code
generator to target different configurations without any change to the formal
descriptions of the algorithms.

@@

In this section we present a prototype implementation and evaluate its
performance. We first present the implementation consisting of the hardware
platform, network and software stack. We first take an analytical approach and
look at the handling of common data and the size of the source code. Next, we
use the implementation to quantify its merits using a side-by-side comparison
of generated code versus a manual implementation. We consider experimental
results regarding usage of the network, execution time of the event-loop and
the image size.

@@

Looking at the generated source code is a straightforward way to validate the
framework. A simple metric is comparing the amount of code versus the size of
the formal description of ID algorithms. This can be an indication of the
development effort and complexity. In this experiment, implementing both
algorithms manually results in 637 lines of C code. The corresponding \NAME
implementation results in 156 lines of DSL code. This is largely thanks to the
higher level coding constructs, as introduced in section \ref{language-support}
dealing with \NAME's language support.

Next we look at data structures. Sharing data eliminates redundant information,
such as a network address, from the scope of each individual algorithm.
Analysis of the source code with both algorithms implemented, shows that this
reduces the memory footprint for one node entry by 2 bytes, from 27 bytes to 25
bytes. In our implementation we used the ZigBee network address, which indeed
consists of 16 bits. This results in a reduction of 7.4\%. Using ZigBee's 64
bits unique addresses, this reduction can easily become 20.5\%.

@@

Besides this analysis we also instrumented the code and evaluated four
criteria: the network usage in number of packets and bytes, the time required
to perform once cycle of the event loop and the image size of the resulting
compiled code. To do this, we collected these metrics for four situations:
without ID algorithms, with a watchdog, with reputation-tracking and with both
algorithms implemented.

@@

Secondly, when turning to the goal to reduce the network usage, we see that
both the number of packets as well as the amount of bytes sent have dropped by
13\% and 7\%.

@@

@@

Adding multiple intrusion detection algorithms to wireless resource-constrained
devices requires fusion of algorithms on a functional level. This can reduce
execution inefficiencies as well as redundant memory usage and network access.

@@

As a next step, we will implement more ID algorithms using \NAME and continue
the experiments on a larger scale. This way we will also be able to verify the
initially observed inclination in the evolution of the image size with growing
numbers of algorithms.

@@

This research is partially funded by the Research Fund KU Leuven and partially
by the EU FP7 project NESSoS\@.
